ODGI: understanding pangenome graphs

### 2021-10-30 manuscript/main.tex
italianization: implementation; explicit who is depth, use model only for the pangenomes, typos, little changes
italianization: resultsn; avoid acronyms not introduced, simplify, typos

ODGI improves on prior efforts, based on issues that arose during our work with high-quality \textit{de novo} assemblies that cover almost all parts of the human genome~\citep{Logsdon_2021,Nurk_2021}.
For example, we find that it is necessary to support graphs that have regions with paths that present very high depth nodes.
Such motifs can occur in collapsed repeat structures generated by ambiguous sequence homology relationships in repeats found in the centromeres and other segmental duplications.
If we cannot process such regions, we have only two options: 1) we can remove these regions, or 2) we can leave them unaligned.
However, neither of these solutions allows us to investigate their biological features.
To seamlessly represent such difficult regions, we follow an approach implemented in the dynamic version of the Graph BWT (GBWT)~\citep{Siren:2020} and built a node-centric, dynamic, compressed model of the paths.
This design supports node-local modification and update of the graph, which lets us operate on paths in parallel.

@@

\begin{comment}
%\subsection{Core functionality}

In the variation graph model, paths have to respect the graph’s topology: this can be verified with odgi validate, to ensure no errors in the input or edited graphs.
In variation graphs the coordinates are provided by the embedded path sequences.
Indeed, the node IDs are not meant to be stable. odgi position finds, translates, and liftovers graph and path positions between different graphs by exploiting their shared path sequences (Figure 1.B).
\end{comment}

@@

ODGI provides a set of first-order interrogative and manipulative operations on pangenome graphs.
We have established these tools to support our exploration of graphs built from hundreds of large eukaryotic genomes.
They are practical and able to work with high levels of graph complexity, even with regions of very high path depth (10$^5$ to 10$^6$-fold coverage).
ODGI's tools cover common operations that we have found to be essential when working with these complex biological models:

\begin{itemize}
\item \textit{odgi build} constructs the ODGI data model from GFA (\S\ref{sec:build}).
\item \textit{odgi view} converts the graph into standard text formats (\S\ref{sec:text}).
\item \textit{odgi matrix} derives the pangenome matrix (\S\ref{sec:text}).
\item \textit{odgi paths} lists and extracts paths in FASTA (\S\ref{sec:text}).
\item \textit{odgi flatten} converts the graph to FASTA and BED (\S\ref{sec:text}).
\item \textit{odgi viz} provides a scalable linear visualization of the graph (\S\ref{sec:viz}).
\item \textit{odgi draw} renders a 2D image of the graph (\S\ref{sec:viz}).
\item \textit{odgi extract} subsets the graph based on path ranges (\S\ref{sec:extract}).
\item \textit{odgi chop} breaks long nodes into shorter ones (\S\ref{sec:edit}).
\item \textit{odgi unchop} combines redundant nodes (\S\ref{sec:edit}).
\item \textit{odgi explode} breaks the graph into connected components (\S\ref{sec:edit}).
\item \textit{odgi squeeze} unifies disjoint graphs (\S\ref{sec:edit}).
\item \textit{odgi prune} removes complex regions (\S\ref{sec:edit}).
\item \textit{odgi position} lift coordinates between paths and graph positions (\S\ref{sec:untangle}).
\item \textit{odgi untangle} deconvolutes paths relative to a reference (\S\ref{sec:untangle}).
\item \textit{odgi tips} finds path end points relative to a reference (\S\ref{sec:untangle}).
\item \textit{odgi sort} orders the graph with sorting pipelines (\S\ref{sec:sort}).
\item \textit{odgi layout} establishes a 2D layout using SGD (\S\ref{sec:sort}).
\item \textit{odgi stats} provides numerical properties of the graph (\S\ref{sec:metrics}).
\item \textit{odgi bin} generates a summarized view of the graph (\S\ref{sec:metrics}).
\item \textit{odgi depth} describes path depth over graph and path positions (\S\ref{sec:metrics}).
\item \textit{odgi degree} describes node degree over graph and path positions (\S\ref{sec:metrics}).
\end{itemize}

Each tool focuses on a small set of related operations.
Most read or write the native ODGI format (`og' extension) (Figure \ref{fig:operations}) and work with standard text based data formats common to bioinformatics.
This supports the implementation of flexible, composable, graph processing pipelines based on a common data type.
While the pangenome graph is a first-class entity in ODGI, we focus on the embedded paths to provide a universal coordinate system.
By considering all paths in the graph as potential reference or query sequences, we make the graph invisible to downstream tools that operate on collections of sequences (e.g. BEDTools, SAMtools \citep{Li2009}, aligners).
This approach benefits from the information in the graph without strongly embedding our methods in this difficult new research context.

>>>>

ODGI improves on prior efforts, based on issues that arose during our work with high-quality \textit{de novo} assemblies that cover almost all parts of the human genome~\citep{Logsdon_2021,Nurk_2021}.
For example, we find that it is necessary to support graphs that have regions where paths present very high depth nodes.
Such motifs can occur in collapsed repeat structures generated by ambiguous sequence homology relationships in repeats found in the centromeres and other segmental duplications.
If we cannot process such regions, we have only two options: 1) we can remove these regions, or 2) we can leave them unaligned.
However, neither of these solutions allows us to investigate their biological features.
To seamlessly represent such difficult regions, we follow an approach implemented in the dynamic version of the Graph BWT (GBWT)~\citep{Siren:2020} and built a node-centric, dynamic, compressed model of the paths.
This design supports node-local modification and update of the graph, which lets us operate on paths in parallel.

@@

%\subsection{Core functionality}
%In the variation graph model, paths have to respect the graph’s topology: this can be verified with odgi validate, to ensure no errors in the input or edited graphs.
%In variation graphs the coordinates are provided by the embedded path sequences.
%Indeed, the node IDs are not meant to be stable. odgi position finds, translates, and liftovers graph and path positions between different graphs by exploiting their shared path sequences (Figure 1.B).

@@

ODGI provides a set of interrogative and manipulative operations on pangenome graphs.
We have established these tools to support our exploration of graphs built from hundreds of large eukaryotic genomes.
ODGI's tools are practical and able to work with high levels of graph complexity, even with regions where paths present very high depth nodes (10$^5$ to 10$^6$-fold depth).

ODGI covers common operations that we have found to be essential when working with complex pangenome graphs:

\begin{itemize}
\item \textit{odgi build} constructs the ODGI data model from GFA (\S\ref{sec:build}).
\item \textit{odgi view} converts the ODGI data model into GFA (\S\ref{sec:text}).
\item \textit{odgi matrix} derives the pangenome matrix (\S\ref{sec:text}).
\item \textit{odgi paths} lists and extracts paths in FASTA (\S\ref{sec:text}).
\item \textit{odgi flatten} converts the graph to FASTA and BED (\S\ref{sec:text}).
\item \textit{odgi viz} provides a linear visualization of the graph (\S\ref{sec:viz}).
\item \textit{odgi draw} renders a 2D image of the graph (\S\ref{sec:viz}).
\item \textit{odgi extract} subsets of the graph based on path ranges (\S\ref{sec:extract}).
\item \textit{odgi chop} breaks long nodes into shorter ones (\S\ref{sec:edit}).
\item \textit{odgi unchop} combines redundant nodes (\S\ref{sec:edit}).
\item \textit{odgi explode} breaks the graph into connected components (\S\ref{sec:edit}).
\item \textit{odgi squeeze} unifies disjoint graphs (\S\ref{sec:edit}).
\item \textit{odgi prune} removes complex regions (\S\ref{sec:edit}).
\item \textit{odgi position} lifts coordinates between paths and graph positions (\S\ref{sec:untangle}).
\item \textit{odgi untangle} deconvolutes paths relative to a reference (\S\ref{sec:untangle}).
\item \textit{odgi tips} finds path end points relative to a reference (\S\ref{sec:untangle}).
\item \textit{odgi sort} orders the graph nodes (\S\ref{sec:sort}).
\item \textit{odgi layout} establishes a 2D layout (\S\ref{sec:sort}).
\item \textit{odgi stats} provides numerical properties of the graph (\S\ref{sec:metrics}).
\item \textit{odgi bin} generates a summarized view of the graph (\S\ref{sec:metrics}).
\item \textit{odgi depth} describes node depth over graph and path positions (\S\ref{sec:metrics}).
\item \textit{odgi degree} describes node degree over graph and path positions (\S\ref{sec:metrics}).
\end{itemize}

Each tool focuses on a small set of related operations.
Most read or write the native ODGI format (`og' extension) (Figure \ref{fig:operations}) and work with standard text based data formats common to bioinformatics.
This supports the implementation of flexible and composable graph processing pipelines based on a common data type.
While the pangenome graph is a first-class entity in ODGI, we focus on the embedded paths to provide a universal coordinate system.
By considering all paths in the graph as potential reference or query sequence, we make the graph invisible to downstream tools that operate on collections of sequences (\textit{e.g.} BEDTools, SAMtools~\citep{Li2009}, aligners).
This approach benefits from the information in the graph without strongly embedding our methods in this difficult new research context.
